Self-reported LLM usage and outcomes on a data science project: Evidence from two undergraduate data science courses

### 2025-04-12 paper/paper.qmd
Minor changes in text of introduction

Like all new tools LLMs have created both excitement and apprehension. ChatGPT's public release on November 30, 2022, brought LLMs into the mainstream conversation. LLMs have quickly been used in both industry and academia. By now many people, especially educators and students, have some experience with LLMs in both personal and professional contexts.

@@

The tasks involved in a trustworthy data science workflow can be generally broken down into a few key competencies [@Adhikari2021Interleaving; @Gibbs2021Building]. One is programming, which is done when cleaning, analyzing, and visualizing data often using programming languages like R or Python. Another is writing, which is primarily done when communicating results. The potential for LLMs to positively affect students' academic performance in data science follows from their already-demonstrated capability in those competencies in adjacent professional fields.

@@

In this paper we are interested in better understanding LLMs as a tool for producing trustworthy data science. We study how they were used by students in two upper-year undergraduate data science courses -- one in Winter 2024 (January to April) and another in Fall 2024 (September to December) -- and whether students who used LLMs tended to have higher scores than those who did not. We focus on the association between student academic performance and their LLM usage. Specifically, we examine the relationship between the mark a student got on a final paper and self-reported measures of student LLM usage, as well as student attitudes toward LLMs in general. This is based on students' final papers and a survey, conducted in third-year undergraduate data science courses at the University of REDACTED. By examining how students interact with and perceive LLMs as tools, and how these variables translate into student outcomes, current practice with regard to LLM integration in data science can be better understood, leading to better recommendations for their future development.

We find some mild association between the mark that a respondent received and their self-reported LLM usage. And this is especially the case for respondents that were not native English speakers. More concretely, there was a considerable increase in the number of respondents who self-reported extensive LLM usage and also those who considered LLM usage to be appropriate and ethical for university, when we compared the results in the class at the start of 2024 with the class at the end of 2024. Taken together it is clear that considerable additional work is needed understanding the appropriate use of generative AI in statistics and data science pedagogy.

@@

To investigate students' usage and attitudes towards LLMs and how they related to their academic performance, a dataset containing their usage and attitudes, coursework, and academic performance was constructed. This was based on three components:

1. an optional survey;
2. self-reported LLM usage; and
3. student marks on their final paper.

@@

In STA304, there were 183 responses to the survey. Of these, 153 respondents provided authorization for their data to be collected and used. Three of those respondents submitted the survey twice and their second response was removed. One student submitted their final paper too late to be of use, and so their survey was removed. Nine respondents did not include a statement on LLM usage in the README of the GitHub repository of their final paper. One respondent did not provide a usable GPA response. Finally, one respondent failed an attention check question halfway through the survey "Please select the option blue".

@@

Students from a range of years took both courses, but most respondents were in their 3rd or 4th year of study (@tbl-year-dist). The prerequisite two-course sequence is typically completed by students in their second year, would make it difficult to take either course earlier than their 3rd year.

>>>>

Like all new tools LLMs have created both excitement and apprehension. ChatGPT's public release on November 30, 2022, brought LLMs into the mainstream conversation. LLMs have rapidly gained traction in both industry and academia. By now many people, especially educators and students, have some experience with LLMs in both personal and professional contexts.

@@

The tasks involved in a trustworthy data science workflow can be generally broken down into a few key competencies [@Adhikari2021Interleaving; @Gibbs2021Building]. One is programming, which is done when cleaning, analyzing, and visualizing data often using programming languages like R or Python. Another is writing, which is primarily done when communicating results. The potential for LLMs to positively affect students' academic performance in data science stems from the capabilities these models have already demonstrated in adjacent professional fields.

@@

In this paper we are interested in better understanding LLMs as a tool for producing trustworthy data science. We study how they were used by students in two upper-year undergraduate data science courses -- one in Winter 2024 (January to April) and another in Fall 2024 (September to December) -- and whether students who used LLMs tended to have higher scores than those who did not. We focus on the association between student academic performance and their LLM usage. Specifically, we examine the relationship between self-reported measures of student LLM usage and the mark a student got on a final paper, as well as student attitudes toward LLMs in general. This is based on students' final papers and a survey, conducted in third-year undergraduate data science courses at the University of REDACTED. By examining how students interact with and perceive LLMs as tools, and how these variables translate into student outcomes, current practice with regard to LLM integration in data science can be better understood, leading to better recommendations for their future development.

We find some mild association between self-reported LLM usage and the mark received on the final paper, especially among students who were not native English speakers. More concretely, there was a considerable increase in the number of respondents who self-reported extensive LLM usage and also those who considered LLM usage to be appropriate and ethical for university, when we compared the results in the class at the start of 2024 with the class at the end of 2024. Taken together it is clear that considerable additional work is needed understanding the appropriate use of generative AI in statistics and data science pedagogy.

@@

To investigate the relationship between students' LLM usage and attitudes and their academic performance, a dataset containing self-reported LLM usage, attitudes towards LLMs, final paper grades, and overall academic performance was constructed. This was based on three components:

1. an optional survey administered at the end of the course;
2. self-reported LLM usage on the final paper; and
3. student marks on their final paper.

@@

In STA304, there were 183 responses to the survey. Of these, 153 respondents provided authorization for their data to be collected and used. Three of those respondents submitted the survey twice and their second response was removed. One student submitted their final paper too late to be of use, and so their survey was removed. Nine respondents did not include a statement on LLM usage in the README of the GitHub repository of their final paper. One respondent did not provide a usable GPA response. Finally, one respondent failed an attention check question halfway through the survey "Please select the option blue". This left 138 usable responses.

@@

Students from a range of years took both courses, but most respondents were in their 3rd or 4th year of study (@tbl-year-dist). The prerequisite two-course sequence is typically completed by students in their second year, which would make it difficult to take either course earlier than their 3rd year.
